---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data

```{r loading}
library(utils)
unzip("activity.zip")
activity <- read.csv("activity.csv")
library(dplyr)
activity <- mutate(activity, hour_min = formatC(interval,digits = 0,width = 4,flag = "0"))
activity <- mutate(activity, dateTime = strptime(paste(date,hour_min), "%Y-%m-%d %H%M"))
str(activity)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps per day
```{r steps_per_day}
steps_per_day <- activity %>%
    group_by(as.factor(date)) %>%
    summarize(total=sum(steps, na.rm = TRUE)) 
head(steps_per_day)
```

2. Make a histogram of the total number of steps per day
```{r histogram}
library(ggplot2)
ggplot(steps_per_day, aes(x=total)) +
    geom_histogram() +
    labs(x= "Steps per day",
      title = "Steps per day histogram")
```

3. Calculate and report the mean and median of the total number of steps per day
Mean:

```{r mean_steps_day}
mean_steps_day <- mean(steps_per_day$total)
mean_steps_day
```
Median:
```{r median_steps_day}
median_steps_day <- median(steps_per_day$total)
median_steps_day
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r daily_activity}
daily_activity <- activity %>%
    group_by(interval) %>%
    summarize(mean_steps_5min=mean(steps, na.rm = TRUE))
ggplot(daily_activity, aes(x=interval,y=mean_steps_5min))+
    geom_line() +
        labs(x= "5-minute interval", y="average steps across all days",
      title = "Average Daily Pattern")

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

* The maximum steps in a 5-minute interval are:
```{r max_steps}
max_steps <- max(daily_activity$mean_steps_5min)
max_steps
```

* The 5-minute interval containing the max number of steps is:
```{r max_interval}
interval_max <- filter(daily_activity, daily_activity$mean_steps_5min == max(mean_steps_5min))
interval_max
```
That is 8:35 am

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r nas_presence}
na_rows <- sum(is.na(activity$steps))
na_rows
```
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

The strategy I devised is to fill each NA with the average for each interval calculated across all days:
```{r replace_nas}
with_nas <- activity$steps
without_nas <- rep(daily_activity$mean_steps_5min,61)
filled_steps <- coalesce(with_nas,without_nas)

```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r dataset_without_nas}
activity$steps_full <- filled_steps
new_activity_dataset <-select(activity,steps_full,date, interval)
head(new_activity_dataset)
```
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r histogram_without_nas}
steps_per_day2 <- activity %>%
    group_by(as.factor(date)) %>%
    summarize(total=sum(steps_full, na.rm = TRUE)) 
ggplot(steps_per_day2, aes(x=total)) +
    geom_histogram() +
    labs(x= "Steps per day",
      title = "Steps per day histogram (without NAs)")
```
Mean (without NAs):

```{r mean_steps_day2}
mean_steps_day2 <- mean(steps_per_day2$total)
mean_steps_day2
```
Median (without NAs):
```{r median_steps_day2}
median_steps_day2 <- median(steps_per_day2$total)
median_steps_day2
```
Differences on mean:
```{r mean_dif}
mean_steps_day2 - mean_steps_day

```
Differences on median:
```{r median_diff}
median_steps_day2 - median_steps_day
```
Both the mean and the median have increased when replaced NA values.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r weekday_patterns}
new_activity_dataset <- mutate(new_activity_dataset, dates = as.Date(date, "%Y-%m-%d"))
new_activity_dataset <- mutate(new_activity_dataset, weekday=weekdays(new_activity_dataset$dates))
new_activity_dataset$weekend_day <- ifelse(new_activity_dataset$weekday %in% c("Sunday","Saturday"), "weekend", "weekday")
head(new_activity_dataset)
unique(new_activity_dataset$weekend_day)

```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r plot_weekend_vs_weekday}
daily_activity2 <- new_activity_dataset %>%
    group_by(interval, weekend_day) %>%
    summarize(mean_steps_5min=mean(steps_full, na.rm = TRUE))
ggplot(daily_activity2, aes(x=interval,y=mean_steps_5min))+
    geom_line() +
        labs(x= "5-minute interval", y="average steps",
      title = "Daily Pattern: weekdays vs. weekends") + 
    facet_wrap(~weekend_day,ncol = 1)

```

On weekends the activity in he morning is lower and starts later in the day/





